---
title: "Introduction to Health Association Analysis for Movement Behaviour Data"
output: 
  rmarkdown::html_vignette:
    number_sections: true
vignette: >
  %\VignetteIndexEntry{Introduction to Health Association Analysis for Movement Behaviour Data}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

In this tutorial, we will go through a basic health association analysis using movement behaviour data.

As well as external packages, we'll use some helper functions we've prepared previously. If you want to modify them, you can find them on the reference page. 

#  Load in data¶
First we will load required packages, and the data we already prepared.

```{r load-packages}
# First we need to install packages that aren't already present. 
if (!require("pacman")) install.packages("pacman")
pacman::p_load(devtools, ggtern, ggplot2, reshape2, data.table)
library(week2DataChallenge)

# We then load the data prepared in the last tutorial. 
dAll <- data.frame(fread("J:/data_and_prep_utilities/59070_application/ukb-new-hes-feb.csv")) #ukb-data-challenge.csv"))# Substitute with the correct path! 
```

#  Describe and explore the data
## Tables to describe the data
The following code constructs a table to describe accelerometer measured attributes by conventional groupings (e.g. age, sex, self-rated health, season). It does two-way analysis of variance, adjusted for the variables listed as confounder columns.

You might think more about the distribution of some of the accelerometry-derived variables - for some, ANOVA may not be the most appropriate way to test for difference. But it's good for a first description!

TROUBLESHOOTING: if you change some of the confounders you use, and you have a categorical variable with two categories, you need to add dAll$category <- as.factor(dAll$category) to the code somewhere. Otherwise, this code won't cope with the variable.

```{r final-prep}
dAll$sex <- as.factor(dAll$sex)
dAll$ethnicity <- as.factor(dAll$ethnicity)
dAll$smoking <- as.factor(dAll$smoking)
dAll$alcohol <- as.factor(dAll$alcohol)

# FOR REMOVAL
dAll$age_entry_years <- dAll$age_entry/365.25
dAll$broadAgeGroup <- cut(dAll$age_entry_years, breaks = c(40, 50, 60, 70, 80 ), labels = c("40-49", "50-59", "60-69", "70-79"))
```

```{r table}
# specify groups (i.e. major row blocks)
groups <-  c("broadAgeGroup", "sex") # Fill this in with groups you want to look at
# specify accelerometer outcomes (i.e. column headings)
outcomeCols <- c( "acc.overall.avg", "MVPA") # Fill this in with a list of outcomes you want to look at

# specify two-way analysis of variance confounder columns
confounderCols <- c("broadAgeGroup", "sex", "ethnicity", "smoking", "alcohol") # Fill this in with what 
#  you want to adjust for when running the ANOVA

# specify output file name
outHTML<-'table1.html'

# write table header code
html <- '<html><head><title>Acc summary table</title></head><table>'
html <- paste(html, '<tr><th></th>')
html <- paste(html, '<th>Individuals</th>')
for(col  in outcomeCols){
    html<-paste(html, '<th>', col, '</th>')
}
html <- paste(html, '</tr>')
html <- paste(html, '<tr><th></th>')
html <- paste(html, '<th>[ n ]</th>')
html <- paste(html, '<th>[mean &plusmn stdev m<i>g</i>]</th>')
html <- paste(html, '<th>[mean &plusmn stdev MVPA hrs/dy]</th>')

# iterate through each major row block
for(group in groups){
    html<-paste(html, '\n<tr><td><b>', group, '</b></td><td></td><td></td><td>')
    html<-paste(html, '</td><td></td><td></td><td></td><td></td><td></td><td></td></tr>')
    group.tmp <- data.frame(matrix(ncol=4, nrow=0)) #df to store outputs for subsequent effect size calculations
    colnames(group.tmp) <- c("item", "outcome", "avg", "stdev")
    # report mean/stdev/n summary stats for each "item" in group e.g. each age categor
    for(item in levels(dAll[[group]])){
        html<-paste(html, '<tr>')
        html<-paste(html, '<td>', item, '</td>')
        # temp store of all relevant outcome data related to this "item"
        itemData <- dAll[dAll[[group]]==item, outcomeCols]
        n <- nrow(itemData)
        html <- paste(html, '<td align="center">', format(n,big.mark=","), '</td>')

        # iterate through each acceleroemter outcome (i.e. each column)
        for(outcome in outcomeCols){
            html <- paste(html, summary4tableHTML_simple(itemData[[outcome]],1))
            avg <- mean(itemData[[outcome]], na.rm=TRUE)
            stdev <- sd(itemData[[outcome]], na.rm=TRUE)
            group.tmp <- rbind(group.tmp, data.frame(item= item, outcome=outcome,
                                     avg=avg, stdev=stdev))
        }
        html<-paste(html, '</tr>')
    }


# now calculate two-way ANOVA p-value between subgroup "items" for each acc outcome
html<-paste(html, '<tr><td>p value<sup>A</sup></td><td></td>')
for(outcome in outcomeCols){
  tmpconfounderCols <- confounderCols
  if (!(group %in% tmpconfounderCols)){
     tmpconfounderCols <- c(group, tmpconfounderCols)
  }
  formulaStr <- paste(c(outcome,' ~ ', paste(tmpconfounderCols, collapse=' + ')), collapse='')
  itemANOVA <- aov(as.formula(formulaStr), data=dAll)
  # extract p-value for specific variable of interest
  
  tmp <- as.data.frame(summary(itemANOVA)[[1]])
  groupResult <- tmp[grep(group, rownames(tmp)), "Pr(>F)"]
   # write formatted p-value to html
  
  html <- paste(html, pValueHTML_cell(groupResult,1e-300))        
 }
 html <- paste(html, '</tr>')
 
 # calculate effect size between subgroup "items" for each acc outcome
 html <- paste(html, "<tr><td>Cohen's d</td><td></td>")
 for(outcome in outcomeCols){
   # read relevant subgroup mean/stdev values stored from summary stats derived earlier
  grouping <-group.tmp[(group.tmp[, "outcome"] ==outcome),]
  minVal <- min(grouping['avg'])
  maxVal <- max(grouping['avg'])
  lowest <- grouping[(grouping$avg==minVal), c('avg','stdev')]
  highest <- grouping[(grouping$avg==maxVal), c('avg','stdev')]
  
  # calculate Cohen's d effect size and format for HTML output
  cohenD <- effectSize(lowest$avg,lowest$stdev,highest$avg,highest$stdev)
  cohenD <- format(round(cohenD, 2), nsmall=2)
  cohenD <- sub('^(-)?0[.]', '\\1.', cohenD)
  html <- paste(html, '<td align="center">', cohenD, '</td>')
 }

  html <- paste(html, '</tr>')

}
#finalise HTML table
html <- paste(html, '\n</table>')
html <- paste(html, '<br><i><sup>A</sup> Age, sex:</i> ')
html <- paste(html, 'Two-way analysis of variance test used to compare metrics between groups')
html <- paste(html, 'adjusting for age, sex, ethnicity, smoking status and alcohol consumption frequency.')
html <- paste(html, '</html>')
# and write html to file
write(html,file = outHTML)

print(html)
print('standard table done, written to: ', outHTML)

```


## Plots to describe the data
We'll also plot some of the variables to get a feel for how they're distributed.

```{r}
#accelerometer measured physical activity
hist(dAll$acc.overall.avg, breaks=1000, xlim=c(0,100))

#look at deciles
quantile(dAll$acc.overall.avg, prob = c(0.10, 0.50, 0.90), na.rm = TRUE)
```

Now let's look descriptively at something we might expect to vary as activity status varies: BMI.

```{r}
plotVarAndQuintile(dAll, # dataset
                   'acc.overall.avg', # exposure
                   'BMI', # outcome
                   FALSE # [TRUE/FALSE] save plots to PDF
                  )
```

Let's have a look at some of the machine-learned variables, to see if we believe them! Here we'll look at plots of probability of being in a particular behaviour by time of day: 
```{r, eval = FALSE}
plotAverageDay(dAll, "sleep.hourOfDay.", ".avg", "Sleep (probability)")

```
You might want to extend this function to plot the different profiles of different groups (e.g. healthy/ unhealthy individuals, or age groups).

# Run a simple health association analysis
Let's run a minimally (age and sex) adjusted linear model for BMI, against fifths of average acceleration vector magnitude. This is attempting to model statistically the association we were looking at descriptively in the end of the last section.

```{r}
dAll$acc_quintiles <- cut(dAll$acc.overall.avg, breaks = c(quantile(dAll$acc.overall.avg, probs = seq(0, 1, by = 0.2), na.rm = TRUE))) # cut at quintiles
min_adj_lm_BMI <- lm(BMI ~acc_quintiles + broadAgeGroup+ sex, dAll)
```

We can look at the model summary:
```{r}
summary(min_adj_lm_BMI)
```

We can of course look at model diagnostics as usual:
```{r}
plot(min_adj_lm_BMI)
```
The problem with the Q-Q plot is caused by the BMI distribution - log-transforming the outcome would be a good option in this case.  

## Next steps for modelling 

Here we've looked at a very simple linear model. Next steps might be:

- adjusting for more confounders
- looking at different outcomes
- logistic regression for prevalent disease (use the `glm` function)
- Cox regression for incident disease (see for example the `survival` package and the `coxph` function, )

@Aiden - I can include an example of Cox regression too? Or project specific examples? 
